# Pillar Theory: Particles as Standing Horizons

A core prediction of **Pillar Theory** — a monistic unification of quantum mechanics and general relativity — is that elementary particles are not fundamental points or strings, but **stable, self-sustaining horizons** that emerge when the universal energy-quanta substrate (Pillar 1) is forced into Planck-scale localization.

This document presents the complete argument, derivation, and citations as of November 2025.

## 1. Physical Motivation

Attempting to confine the ubiquitous energy-quanta substrate below the Planck length  
$$ l_p = \sqrt{\frac{\hbar G}{c^3}} \approx 1.616 \times 10^{-35} \, \text{m} $$  
requires energy  
$$ E \gtrsim E_p = \sqrt{\frac{\hbar c^5}{G}} \approx 1.22 \times 10^{19} \, \text{GeV} $$  
and produces density approaching  
$$ \rho_p = \frac{c^5}{\hbar G^2} \approx 5 \times 10^{96} \, \text{kg/m}^3. $$

At this threshold the substrate can no longer remain homogeneous: a dynamical boundary (horizon) forms that holographically encodes all internal degrees of freedom on its surface, exactly as proposed by 't Hooft (1993) and Susskind (1995) for black holes.

→ There are no true point particles and no singularities — only horizons.

## 2. Stability via Topological Locking (Pillar 3)

Transient horizons evaporate rapidly (Hawking-like radiation from chaotic vector flows). Long-lived particles correspond to horizons whose internal quanta settle into **coherent, resonant vector loops** — topologically protected configurations (toroidal, figure-8, etc.) that minimise free energy.

Unstable → evaporates in ≈ Planck time  
Stable   → persists for 10^{34} years (proton-like)

## 3. Effective Field Theory Description

We model local density deviation ϕ from the vacuum expectation value using the scalar Lagrangian  
$$ \mathcal{L} = \frac{1}{2} \partial_\mu \phi \partial^\mu \phi - V(\phi) $$  
with the Mexican-hat potential  
$$ V(\phi) = -\frac{\mu^2}{2} \phi^2 + \frac{\lambda}{4} \phi^4 + V_0. $$

The vacuum manifold lies at  
$$ \phi_0 = \sqrt{\frac{\mu^2}{\lambda}} \propto \sqrt{\rho_p} \quad (\text{tied to substrate density}). $$

Spontaneous symmetry breaking yields:
- Massive radial mode → particle mass $$ m \propto \sqrt{\lambda} \, \phi_0 $$
- Massless Goldstone modes → internal phase degrees of freedom encoded on the horizon

The equation of motion in the broken phase (small oscillations around ϕ₀) becomes harmonic with  
$$ \omega^2 = 2\mu^2 = 2\lambda \phi_0^2 > 0, $$  
providing the “locking” frequency that stabilises the horizon radius.

## 4. Thermodynamics of Particle-Horizons

Every stable horizon obeys the Bekenstein–Hawking entropy  
$$ S = \frac{A}{4 l_p^2}, \quad A = 4\pi r_s^2, \quad r_s \approx \frac{2Gm}{c^2}. $$

Temperature and evaporation rates follow directly from Hawking (1975), now applied to elementary particles.

## 5. References

- Bekenstein, J. D. (1973). Black Holes and Entropy. *Phys. Rev. D* **7**, 2333.  
- Hawking, S. W. (1975). Particle Creation by Black Holes. *Commun. Math. Phys.* **43**, 199.  
- 't Hooft, G. (1993). Dimensional Reduction in Quantum Gravity. arXiv:gr-qc/9310026.  
- Susskind, L. (1995). The World as a Hologram. *J. Math. Phys.* **36**, 6377.  
- Higgs, P. W. (1964). Broken Symmetries and the Masses of Gauge Bosons. *Phys. Lett.* **12**, 132.  
- Coleman, S. & Weinberg, E. (1973). Radiative Corrections as the Origin of Spontaneous Symmetry Breaking. *Phys. Rev. D* **7**, 1888.

## 6. To-do / Open Questions (WIP)

- Derive μ² and λ directly from quanta interaction graph statistics  
- Simulate toroidal vector-loop stability thresholds (Python / C++)  
- Compute explicit particle mass spectrum for first three generations  
- Predict modified Hawking evaporation rate for microscopic horizons

---
*Part of Pillar Theory — © Chase Holmes 2025 | MIT License*
